Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-06-16 paper/main.tex
addressed PCF suggestions from slack message (6/15/23 pinned item)

\documentclass[preprint,12pt,a4paper]{elsarticle}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{float}
\restylefloat{table}
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage{placeins}
\usepackage{geometry}
\geometry{left=1in,right=1in,top=1in,bottom=1in,headsep=0pt}

@@

\title{\texttt{davos}: a Python package ``smuggler'' for constructing
  lightweight reproducible notebooks}
\author{Paxton C. Fitzpatrick}
\author{Jeremy R. Manning\corref{cor}}
\ead{Jeremy.R.Manning@Dartmouth.edu}
\cortext[cor]{Corresponding author}
\address{Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755}

\begin{abstract}
  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the
  complexity and time cost of sharing or engaging with reproducible
  science. Here, we propose a lighter-weight solution: the
  \texttt{davos} package.  When used in combination with a
  notebook-based Python project, \texttt{davos} provides a mechanism
  for specifying (and automatically installing) the correct versions
  of the project's dependencies.  The \texttt{davos} package further
  ensures that those packages and specific versions are used every
  time the notebook's code is executed.  This enables researchers to
  share a complete reproducible copy of their code within a single
  Jupyter notebook file.
\end{abstract}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external packages, but
  those packages must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those packages were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} package} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external packages (at their
  correct versions, when specified).  \textbf{Virtual environments}
  allow users to install an isolated copy of Python and all required
  dependencies. This typically entails distributing a configuration
  file (e.g., a \texttt{pyproject.toml}~\cite{CannEtal16} or
  \texttt{environment.yml} file) that specifies all project
  dependencies (including version numbers of external packages)
  alongside the primary code base. Users can then install a
  third-party tool~\cite[e.g.,][]{Anac12, Eust19} to read the file and
  build the environment.  \textbf{Containers} provide a means of
  defining an isolated environment that includes a complete operating
  system (independent of the user's operating system), in addition to
  (optionally) specifying a virtual environment or other
  configurations needed to provide the necessary computing
  environment.  Containers are typically defined using specification
  files (e.g., a plain-text \texttt{Dockerfile}) that instruct the
  virtualization engine regarding how to build the containerized
  environment.  \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary image files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing individual files,
  with no additional installation or configuration needed to run the
  target code.  Systems to the right of the red vertical line support
  precise control over dependencies and versioning.  Notebooks
  enhanced using the \texttt{davos} package are easily shareable and
  require minimal setup costs, while also facilitating high
  reproducibility by enabling precise control over project
  dependencies.}
\label{fig:code-sharing}
\end{figure}

@@

We designed \texttt{davos} to occupy a ``sweet spot'' between these extremes.
\texttt{davos} is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
\texttt{davos}-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run standard Jupyter
notebooks. And like virtual environments, \texttt{davos} provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

The \texttt{davos} package is named after Davos Seaworth, a smuggler
referred to as ``the Onion Knight" from the series \textit{A Song of
  Ice and Fire} by George R. R. Martin~\cite{Mart98}.  The
\texttt{smuggle} keyword provided by \texttt{davos}
is a play on Python's \texttt{import} keyword: whereas importing can
load a package into the Python workspace within the existing rules
and frameworks provided by the Python language, ``smuggling'' provides
an alternative that expands the scope and reach of ``importing.''
Like the character Davos Seaworth (who became famous for smuggling
onions through a blockade on his homeland), we use ``onion'' comments
to precisely control how packages are smuggled into the Python
workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{davos} package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring \texttt{davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The \texttt{davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), and a runtime interface for configuring
\texttt{davos}'s behavior (see Sec.~\ref{subsec:config}). However,
certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend \texttt{davos} to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing \texttt{davos} in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
local environment, \texttt{davos} will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the namespace for immediate use.

Installing new packages in a notebook using standard approaches (e.g., system
commmands) affect the runtime environment. This could lead to undesired
behaviors. For example, running a notebook that installs new packages in the
user's primary system environment might alter their main system installation
and/or containerized environment in unexpected ways (e.g., changing package
versions, causing conflicts with other packages, etc.). To protect against
undesired changes to the runtime environment, \texttt{davos} incorporates its
own virtual environment for managing packages it installs. When \texttt{davos}
is imported, a new virtual environment (folder) is created automatically. The
folder's name may be customized to support multi-notebook projects. Any
\texttt{smuggle}d packages that were not available in the notebook's runtime
environment are installed to the current project folder. The runtime
environment remains unaffected by \texttt{davos}'s behavior (see
Sec.~\ref{subsec:project}).

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion comment.'' An
onion comment is a special type of inline comment that may be placed on a line
containing a \texttt{smuggle} statement to customize how \texttt{davos}
searches for the smuggled package locally and, if necessary, downloads and
installs it. Onion comments follow a simple format based on the ``type
comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to
make managing packages with \texttt{davos} intuitive and familiar. To construct
an onion comment, users provide the name of the installer program (e.g.,
\texttt{pip}) and the same arguments one would use to manually install the
package as desired via the command line: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1} \end{center} Occasionally,
a package's distribution name (i.e., the name used when installing it) may
differ from its top-level module name (i.e., the name used when importing it).
In such cases, an onion comment may be used to ensure that \texttt{davos}
installs the proper package if it cannot be found locally: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2} \end{center} Because onion
comments may be constructed to specify any aspect of the installer's behavior,
they provide a mechanism for precisely controlling how, where, and when
smuggled packages are installed. Critically, if an onion comment includes a
version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific version
requested, or satisfies the given version constraints. If the smuggled package
exists locally, \texttt{davos} will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), \texttt{davos} will download, install, and
load a suitable version of the package instead: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3} \end{center} Onion
comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.): \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4} \end{center}
\texttt{davos} processes onion comments internally before forwarding arguments
to the installer program. In addition to preventing onion comments from being
used as a vehicle for shell injection attacks, this enables \texttt{davos} to
adapt its behavior based on how particular flags will affect the behavior of
the installer program. For example, if an onion comment contains either the
\texttt{-I}/\texttt{--ignore-installed}, \texttt{-U}/\texttt{--upgrade}, or
\texttt{--force-reinstall} flag, \texttt{davos} will not bother checking for a
local copy of the smuggled package before installing a new one: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5} \end{center} Similarly,
the \texttt{--no-input} flag will temporarily enable \texttt{davos}'s
non-interactive mode (see Sec.~\ref{subsec:config}). Installation flags that
affect the user's system outside of the current notebook are disabled by
default (see Sec.~\ref{subsec:projects}).

, and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

\subsubsection{The \texttt{davos} config object}\label{subsec:config}

The \texttt{davos} config object provides a high-level interface for
controlling various aspects of \texttt{davos}'s behavior. After
importing \texttt{davos}, the \texttt{davos.con\-fig} object (a
singleton) exposes configurable options as attributes that can be
modified, displayed in the notebook, or checked programmatically at runtime
(see Sec.~\ref{sec:illustrative-example} for an illustrative example or
Sec.~\ref{subsec:implementation} for implementation details and
additional information). These include:

\begin{itemize}
\item \texttt{.active}: This attribute controls whether support for \texttt{smuggle}
  statements and onion comments is enabled (\texttt{True}) or
  disabled (\texttt{False}).  When \texttt{davos} is first imported,
  the \texttt{.active} attribute is set to \texttt{True}.

\item \texttt{.auto\_rerun}: This attribute controls how
  \texttt{davos} behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter,
  and the code that generates those objects was changed between the
  previously imported and to-be-smuggled versions.  If this attribute
  is set to \texttt{True}, \texttt{davos} will automatically restart
  the notebook kernel and rerun all code up to (and including) the
  current \texttt{smuggle} statement. If set to \texttt{False} (the default),
  \texttt{davos} will instead issue a warning, pause execution, and
  prompt the user to either restart and rerun the notebook, or
  continue running with the previously imported package version until
  the next time the kernel is restarted manually.  Note that, as of
  this writing, the \texttt{.auto\_rerun} attribute is not supported
  in Google Colaboratory notebooks.

\item \texttt{.confirm\_install}: If set to \texttt{True} (default:
  \texttt{False}), \texttt{davos} will require user confirmation
  before installing a smuggled package that does not yet exist in the
  user's environment.

\item \texttt{.noninteractive}: Setting this attribute to
  \texttt{True} (default: \texttt{False}) enables non-in\-ter\-act\-ive
  mode, in which all user interactions (prompts and dialogues) are
  disabled. Note that in non-interactive mode, the
  \texttt{confirm\_install} option is set to \texttt{False}.  If
  \texttt{auto\_rerun} is set to \texttt{False} while in non-interactive
  mode, \texttt{davos} will raise an exception if a smuggled package
  cannot be reloaded, rather than prompting the user.

@@

\item \texttt{.suppress\_stdout}: If this attribute is set to
  \texttt{True} (default: \texttt{False}), \texttt{davos} suppresses
  printed (console) outputs from both itself and the installer program.
  This can be useful when smuggling packages that need to install many
  dependencies and/or generate extensive output. However, if the installer
  program throws an error, both its stdout and stderr streams will be
  displayed alongside the Python traceback to allow for debugging.
\end{itemize}

\noindent The top-level \texttt{davos} namespace also defines
convenience functions for setting and checking whether \texttt{davos}
is active (\texttt{davos.activate()}; \texttt{davos.deactivate()};
\texttt{davos.is\_active()}) as well as the \texttt{davos.configure()}
function, which allows setting multiple configuration options
simultaneously.

Because \texttt{davos} can install new packages, running the code in a
\texttt{davos}-enhanced notebook might (in principle) affect the behavior of
\textit{other} Python-based software (e.g., other notebooks, scripts, etc.) by
altering which packages are installed in the runtime environment. This could
lead to undesired consequences. For example, suppose Person A develops a
notebook (Notebook A) for their research project. We will assume that Notebook
A does not use \texttt{davos} to manage project dependencies.  If Person A
runs a \texttt{davos}-enhanced Notebook B, e.g., sent by another developer,
might this unexpectedly affect the behavior of Notebook A?

We implemented a ``project'' system in \texttt{davos} to protect against the
above scenario. By default, importing \texttt{davos} creates a new project
folder in the user's home directory (contained within a hidden \texttt{.davos}
folder). The default project name is computed to uniquely identify each
notebook according to its filename and path. Any packages that were not
originally available in the notebook's runtime environment are installed to the
notebook's project directory. When external libraries are \texttt{smuggle}d,
\texttt{davos} temporally appends the current project directory to the search
path. Because the user's system path remains unchanged, and because none of the
runtime environment's packages are altered, the user's system and runtime
environment remain unaffected (aside from installing the \texttt{davos} package
itself to the runtime environment).

Finally, if the user \textit{does} wish to modify their runtime environment,
this may be done by setting \texttt{davos.project} to \texttt{None}. Doing so
will cause any packages installed by \texttt{davos} to affect the user's
runtime environment.  This is generally not recommended, as it can lead to
unintended consequences for other code that shares the runtime environment.

@@

Although \texttt{davos} is designed to \textit{appear} to add a new
keyword to Python's vocabulary, this illusion is actually created through
several ``hacks'' that make use of the notebook's IPython backend
for processing and executing users' code.  Specifically, when
\texttt{davos} is first imported, or when it is activated after having been
set to an inactive state, two actions are triggered.  First, the
\texttt{smuggle()} function is injected into the IPython user
namespace.  Second, the \texttt{davos} parser is registered as a
custom IPython input transformer.

>>>>

\documentclass[preprint,12pt,a4paper]{elsarticle}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{float}
\restylefloat{table}
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage{placeins}
\usepackage{geometry}

\geometry{left=1in, right=1in, top=1in, bottom=1in, headsep=0pt}

@@

\title{Davos: a Python package ``smuggler'' for constructing
  lightweight reproducible notebooks}
\author{Paxton C. Fitzpatrick}
\author{Jeremy R. Manning\corref{cor}}
\ead{Jeremy.R.Manning@Dartmouth.edu}
\cortext[cor]{Corresponding author}
\address{Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755}

\begin{abstract}
  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the
  complexity and time cost of sharing or engaging with reproducible
  science. Here, we propose a lighter-weight solution: the
  Davos package.  When used in combination with a
  notebook-based Python project, Davos provides a mechanism
  for specifying (and automatically installing) the correct versions
  of the project's dependencies.  The Davos package further
  ensures that those packages and specific versions are used every
  time the notebook's code is executed.  This enables researchers to
  share a complete reproducible copy of their code within a single
  Jupyter notebook file.
\end{abstract}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external packages, but
  those packages must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those packages were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{Davos package} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external packages (at their
  correct versions, when specified).  \textbf{Virtual environments}
  allow users to install an isolated copy of Python and all required
  dependencies. This typically entails distributing a configuration
  file (e.g., a \texttt{pyproject.toml}~\cite{CannEtal16} or
  \texttt{environment.yml} file) that specifies all project
  dependencies (including version numbers of external packages)
  alongside the primary code base. Users can then install a
  third-party tool~\cite[e.g.,][]{Anac12, Eust19} to read the file and
  build the environment.  \textbf{Containers} provide a means of
  defining an isolated environment that includes a complete operating
  system (independent of the user's operating system), in addition to
  (optionally) specifying a virtual environment or other
  configurations needed to provide the necessary computing
  environment.  Containers are typically defined using specification
  files (e.g., a plain-text \texttt{Dockerfile}) that instruct the
  virtualization engine regarding how to build the containerized
  environment.  \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary image files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing individual files,
  with no additional installation or configuration needed to run the
  target code.  Systems to the right of the red vertical line support
  precise control over dependencies and versioning.  Notebooks
  enhanced using the Davos package are easily shareable and
  require minimal setup costs, while also facilitating high
  reproducibility by enabling precise control over project
  dependencies.}
\label{fig:code-sharing}
\end{figure}

@@

We designed Davos to occupy a ``sweet spot'' between these extremes.
Davos is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
Davos-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run standard Jupyter
notebooks. And like virtual environments, Davos provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

The Davos package is named after Davos Seaworth, a smuggler referred
to as ``the Onion Knight" from the series \textit{A Song of Ice and Fire} by
George R. R. Martin~\cite{Mart98}. The \texttt{smuggle} keyword provided by
Davos is a play on Python's \texttt{import} keyword: whereas importing
can load a package into the Python workspace within the existing rules and
frameworks provided by the Python language, ``smuggling'' provides an
alternative that expands the scope and reach of ``importing.'' Like the
character Davos Seaworth (who became famous for smuggling onions through a
blockade on his homeland), we use ``onion'' comments to precisely control how
packages are smuggled into the Python workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The Davos package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring Davos's behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The Davos package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), and a runtime interface for configuring
Davos's behavior (see Sec.~\ref{subsec:config}). However,
certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
Davos is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, Davos detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend Davos to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing Davos in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
local environment, Davos will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the namespace for immediate use.

Installing new packages in a notebook using standard approaches (e.g., system
commmands) affect the runtime environment. This could lead to undesired
behaviors. For example, running a notebook that installs new packages in the
user's primary system environment might alter their main system installation
and/or containerized environment in unexpected ways (e.g., changing package
versions, causing conflicts with other packages, etc.). To protect against
undesired changes to the runtime environment, Davos incorporates its
own virtual environment for managing packages it installs. When Davos
is imported, a new virtual environment (folder) is created automatically. The
folder's name may be customized to support multi-notebook projects. Any
\texttt{smuggle}d packages that were not available in the notebook's runtime
environment are installed to the current project folder. The runtime
environment remains unaffected by Davos's behavior (see
Sec.~\ref{subsec:projects}).

@@

For greater control over the behavior of \texttt{smuggle} statements, Davos
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how Davos searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with Davos
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center} \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center} Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment may be used to ensure
that Davos installs the proper package if it cannot be found locally:
\begin{center} \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center} Because onion comments may be constructed to specify any aspect of
the installer's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, Davos will ensure that
the version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the smuggled
package exists locally, Davos will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), Davos will download, install, and load a
suitable version of the package instead: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3} \end{center} Onion
comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.): \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4} \end{center} Davos
processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing onion comments from being used as
a vehicle for shell injection attacks, this enables Davos to adapt its behavior
based on how particular flags will affect the behavior of the installer
program. For example, if an onion comment contains either the
\texttt{-I}/\texttt{--ignore-installed}, \texttt{-U}/\texttt{--upgrade}, or
\texttt{--force-reinstall} flag, Davos will not bother checking for a local
copy of the smuggled package before installing a new one: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5} \end{center} Similarly,
the \texttt{--no-input} flag will temporarily enable Davos's non-interactive
mode (see Sec.~\ref{subsec:config}). Installation flags that affect the user's
system outside of the current notebook are disabled by default (see
Sec.~\ref{subsec:projects}), and installing a smuggled package into a custom
directory (\texttt{<dir>}) using the \texttt{--target <dir>} flag will cause
Davos to prepend \texttt{<dir>} to the module search path (i.e.,
\texttt{sys.path}), if necessary, so the package can be imported.

Because Davos can install new packages, running the code in a
Davos-enhanced notebook might (in principle) affect the behavior of
\textit{other} Python-based software (e.g., other notebooks, scripts, etc.) by
altering which packages are installed in the runtime environment. This could
lead to undesired consequences. For example, suppose Person A develops a
notebook (Notebook A) for their research project. We will assume that Notebook
A does not use Davos to manage project dependencies.  If Person A
runs a Davos-enhanced Notebook B, e.g., sent by another developer,
might this unexpectedly affect the behavior of Notebook A?

We implemented a ``project'' system in Davos to protect against the
above scenario. By default, importing Davos creates a new project
folder in the user's home directory (contained within a hidden \texttt{.davos}
folder). The default project name is computed to uniquely identify each
notebook according to its filename and path. Any packages that were not
originally available in the notebook's runtime environment are installed to the
notebook's project directory. When external libraries are \texttt{smuggle}d,
Davos temporally appends the current project directory to the search
path. Because the user's system path remains unchanged, and because none of the
runtime environment's packages are altered, the user's system and runtime
environment remain unaffected (aside from installing the Davos package
itself to the runtime environment).

Finally, if the user \textit{does} wish to modify their runtime environment,
this may be done by setting \texttt{davos.project} to \texttt{None}. Doing so
will cause any packages installed by Davos to affect the user's
runtime environment.  This is generally not recommended, as it can lead to
unintended consequences for other code that shares the runtime environment.

\subsubsection{Configuring and querying Davos}\label{subsec:config}

Davos's behavior may be customized by modifying a set of attributes attached to
the \texttt{davos} module object that is added to the workspace when Davos is
imported. These attributes may be modified, displayed, or checked
programmatically at runtime (see Sec.~\ref{sec:illustrative-example} for an
illustrative example or Sec.~\ref{subsec:implementation} for implementation
details and additional information). These include:

\begin{itemize}
\item \texttt{.active}: This attribute controls whether support for \texttt{smuggle}
  statements and onion comments is enabled (\texttt{True}) or
  disabled (\texttt{False}).  When Davos is first imported,
  the \texttt{.active} attribute is set to \texttt{True}.

\item \texttt{.auto\_rerun}: This attribute controls how
  Davos behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter,
  and the code that generates those objects was changed between the
  previously imported and to-be-smuggled versions.  If this attribute
  is set to \texttt{True}, Davos will automatically restart
  the notebook kernel and rerun all code up to (and including) the
  current \texttt{smuggle} statement. If set to \texttt{False} (the default),
  Davos will instead issue a warning, pause execution, and
  prompt the user to either restart and rerun the notebook, or
  continue running with the previously imported package version until
  the next time the kernel is restarted manually.  Note that, as of
  this writing, the \texttt{.auto\_rerun} attribute is not supported
  in Google Colaboratory notebooks.

\item \texttt{.confirm\_install}: If set to \texttt{True} (default:
  \texttt{False}), Davos will require user confirmation
  before installing a smuggled package that does not yet exist in the
  user's environment.

\item \texttt{.noninteractive}: Setting this attribute to
  \texttt{True} (default: \texttt{False}) enables non-in\-ter\-act\-ive
  mode, in which all user interactions (prompts and dialogues) are
  disabled. Note that in non-interactive mode, the
  \texttt{confirm\_install} option is set to \texttt{False}.  If
  \texttt{auto\_rerun} is set to \texttt{False} while in non-interactive
  mode, Davos will raise an exception if a smuggled package
  cannot be reloaded, rather than prompting the user.

@@

\item \texttt{.suppress\_stdout}: If this attribute is set to
  \texttt{True} (default: \texttt{False}), Davos suppresses
  printed (console) outputs from both itself and the installer program.
  This can be useful when smuggling packages that need to install many
  dependencies and/or generate extensive output. However, if the installer
  program throws an error, both its stdout and stderr streams will be
  displayed alongside the Python traceback to allow for debugging.

\item \texttt{.project}: This attribute is a string that specifies the name of
the ``project'' associated with the current notebook. As described in
Section~\ref{subsec:projects}, a notebook's project determines where and how
any \textttt{smuggle}d dependencies are installed if they are not available in
the current runtime environment. By default, this attribute is named according
to the current notebook's absolute file path. However, the project name may be
customized to enable shared dependency installations across notebooks (see
Sec.~\ref{subsec:projects}).

\end{itemize}

\noindent Davos namespace also defines the \texttt{davos.configure()} function,
which allows setting multiple configuration options simultaneously. In addition
to the above configurable attributes, the \texttt{davos} object also includes
several read-only attributes that contain potentially useful information about
the current environment or Davos's behavior:

\begin{itemize}

\item \texttt{.environment}: This attribute's value is a string describing the notebook
environment Davos was imported into. As of the current version (0.2.0), this
attribute will be set to either \texttt{“IPython<7.0"}, \texttt{“IPython>=7.0”}, or \texttt{“Colaboratory”}.

  \item \texttt{.ipython\_shell}: This attribute contains the global IPython \texttt{InteractiveShell} instance underlying the notebook kernel session.

  \item \texttt{.smuggled}: This attribute is set to a Python dictionary that functions as a cache of any \texttt{smuggle} commands run during the current
session.  The dictionary's keys are package names and the values are arguments passed via the corresponding smuggle statement's onion comment.

  \item \texttt{.all\_projects}: This attribute contains a list of all local projects (i.e., projects with virtual environment directories located in \texttt{\$HOME/.davos/projects}).  See Section~\ref{subsec:projects} for additional information about Davos projects.

\end{itemize}

@@

Although Davos is designed to \textit{appear} to add a new
keyword to Python's vocabulary, this illusion is actually created through
several ``hacks'' that make use of the notebook's IPython backend
for processing and executing users' code.  Specifically, when
Davos is first imported, or when it is activated after having been
set to an inactive state, two actions are triggered.  First, the
\texttt{smuggle()} function is injected into the IPython user
namespace.  Second, the Davos parser is registered as a
custom IPython input transformer.

IPython preprocesses all executed code as plain text before it is sent
to the Python compiler, in order to handle special constructs like
\texttt{\%magic} and \texttt{!shell} commands. Davos uses
this process to transform \texttt{smuggle} statements into
syntactically valid Python code. The Davos parser uses a
regular expression to match lines of code containing \texttt{smuggle}
statements (and, optionally, onion comments), extract relevant
information from their text, and replace them with equivalent calls to
the \texttt{smuggle()} function. For example, if a user runs a
notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
The call to the \texttt{smuggle()} function carries out
Davos's central logic by determining whether the smuggled
package must be installed, carrying out the installation if necessary,
and subsequently loading it into the namespace. This process is
outlined in Figure~\ref{fig:flow-chart}. Because the
\texttt{smuggle()} function is defined in the notebook namespace, it
is also possible (though never necessary) to call it
directly. Deactivating Davos will delete the name
``\texttt{smuggle}'' from the namespace, unless its value has been
overwritten and no longer refers to the \texttt{smuggle()}
function. It will also deregister the Davos parser from the
set of input transformers run when each notebook cell is
executed. While the overhead added by the Davos parser is
minimal, this may be useful, for example, when optimizing or precisely
profiling code.
